Reasonably Programmable Literal Notation

### 2018-06-18 icfp18/syntax-icfp18.tex
edits to 2.0

%% Author with single affiliation.
\author{Cyrus Omar}
\authornote{The majority of this research was performed while the first author was at Carnegie Mellon University.}          %% \authornote is optional;
                                        %% can be repeated if necessary
\affiliation{
  \institution{University of Chicago}            %% \institution is required
  \city{Chicago}
  \state{IL}
  % \country{USA}
}
\email{comar@cs.uchicago.edu}          %% \email is recommended
% \affiliation{
%   \institution{Carnegie Mellon University}           %% \institution is required
%   \city{Pittsburgh}
%   \state{PA}
%   % \country{USA}
% }

@@

Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-6 outline the Camlp4 mechanism \cite{de2003camlp4}, but Sugar*/SugarHaskell is similar \cite{erdweg2011sugarj,erdweg2012layout}. 
Line 8 uses one of these active syntax extensions to construct an encoding of a query in the rather obscure database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer examining the program as presented, and unfamiliar with (i.e. holding abstract) the details elided on Lines 1-6, cannot easily answer questions like the following:

@@

\paragraph{Paper Outline} 
% \begin{itemize}
% \item 
Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs. It also describes experimental integration with MetaOCaml, which can help providers reason about type correctness.
% \item 
Sec.~\ref{sec:sptsms} then briefly introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
 % This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors). 
 Having introduced the basic machinery by example, we proceed in 
%Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies. Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs. 
Sec.~\ref{sec:setlms-formally} to describe a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form. The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.
% Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs for Reason without modifying OCaml's type system. % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct. 
This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page:

@@

Consider the recursive datatype \li{Regex.t} defined by Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
Note that in Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}. Even with the aid of this shorthand, however, constructing regexes in this way is syntactically costly. Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings. 
We solve this problem in Fig.~\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings. Fig.~\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex. This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}. However, this approach breaks down when one needs to construct regexes compositionally. For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases. However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}. These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}. This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation. When some of these strings come from untrusted sources, e.g. web forms as is common in bioinformatics applications, the result can be a breach in privacy.

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  }
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer;
      parser RegexParser.start;
      dependencies
        { module Regex = Regex; }
    }
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
\label{fig:regex-tlm-def}
\end{subfigure}

>>>>

%% Author with single affiliation.
\author{Cyrus Omar}
\authornote{The majority of this research was performed while the first author attended Carnegie Mellon University.}          %% \authornote is optional;
                                        %% can be repeated if necessary
\affiliation{
  \institution{University of Chicago}            %% \institution is required
  \city{Chicago}
  \state{IL}
  % \country{USA}
}
\email{comar@cs.uchicago.edu}          %% \email is recommended
% \affiliation{
%   \institution{Carnegie Mellon University}           %% \institution is required
%   \city{Pittsburgh}
%   \state{PA}
%   % \country{USA}
% }

@@

Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig.~\ref{fig:K-dialect}, which is written in an emerging dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-6 outline the Camlp4 mechanism \cite{de2003camlp4}; Sugar*/SugarHaskell is similar \cite{erdweg2011sugarj,erdweg2012layout}. 
Line 8 uses one of these active syntax extensions to construct an encoding of a query in the rather obscure database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer examining the program as presented, and unfamiliar with (i.e. holding abstract) the details elided on Lines 1-6, cannot easily answer questions like the following:

@@

\paragraph{Paper Outline} 
% \begin{itemize}
% \item 
Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs. It also describes experimental integration with MetaOCaml, which can help providers reason about type correctness.
% \item 
Sec.~\ref{sec:sptsms} then briefly introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
 % This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors). 
 Having introduced the basic machinery by example, we proceed in 
%Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies. Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs. 
Sec.~\ref{sec:setlms-formally} to describe a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form. The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.
% Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs for Reason without modifying OCaml's type system. % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct. 
This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the \li{Relit} project page:

@@

Consider the recursive datatype \li{Regex.t} defined in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
Note that in Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}. Even with the aid of this shorthand, however, constructing regexes in this way is syntactically costly. Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings. 
We solve this problem in Fig.~\ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes (delimited by \li{$(} and \li{)}) and strings (delimited by \li{$$(} and \li{)}). Fig.~\ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied.
% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex. This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}. However, this approach breaks down when one needs to construct regexes compositionally. For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases. However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}. These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}. This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation. When some of these strings come from untrusted sources, e.g. web forms as is common in bioinformatics applications, the result can be a breach in privacy.

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  }
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    }
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
\label{fig:regex-tlm-def}
\end{subfigure}
